---
title: "An example Rmarkdown post"
authors:
  - Rohit Ram
date: "2020-11-12"
output: html_document
bibliography: ["casual.bib"]
link-citations: true
categories: [blogpost]
commentable: true
---


## This shows how to use `Rmarkdown` in your blogs.



```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = T, message = FALSE, warning = F)
library(ggplot2)
library(dplyr)
library(tufte)
library(patchwork)
library(ggExtra)
library(dagitty)
library(ggdag)
library(data.table)
```

Most statistics students will be familiar with the phrase "correlation isn't causation," however, this doesn't feature strongly in the remainder of their educations. To overcome this hurdle, the researchers' best practice in experimental design is the randomized controlled trial. However, there are only specific experiments that we can perform. For example, to test the whether smoking causes cancer, we can't force subjects to smoke.  In general, restricting ourselves to experimental studies to determine causation is incredibly limiting (especially for data scientists). We want to make the same causal conclusions from observational studies and those from experimental studies. We can do that by studying causal inference.

## Simpson's Paradox {#simpsons}
An example of the importance of understanding causal relationships is given by Simpson's Paradox [@simpson1951interpretation], which describes a peculiar phenomenon that can present in data sets, where a correlation between two variables is present in one direction but reverses in each stratum of the data. The paradox expressed best through an example:

```{r simpsons-before, echo=F, fig.cap="The results of an experiment, where x-axis represents how much exercise an individual does in hours, and y-axis represents cholestral measurment for the same individual."}
set.seed(123)

n <- 1000
# age <- sample(seq(20,70,10), size=n, replace=TRUE) + rnorm(n, mean=5, sd = 2)
age <- rnorm(n, mean=60, sd = 12)
exercise <- (age-45)/13 + rnorm(n, mean=5, sd = 1)
cholesteral <- -4 * exercise + 1 * age + rnorm(n, mean = 170, sd=3)
simpsons_data <- data.table(age = age, exercise = exercise, cholesteral=cholesteral)
simpsons_data %>%
  mutate(age = as.factor(signif(age,1))) %>%
  ggplot(aes(x=exercise, y= cholesteral)) +
  geom_point() +
  geom_smooth(method = 'lm', se=F)+
  labs(x="Exercise [Hours per Week]", y="Cholesterol [Total cholesterol mg/dL]")
```
Figure \@ref(fig:simpsons-before) shows a positive correlation in an experiment that measures individuals' exercise per week and cholesterol. At first glance, this seems absurd, but when we partition the data by another causal variable, this seems reasonable:
```{r simpsons-after, echo=F, fig.cap="The same results as the experiment above, partioned by age"}
simpsons_data %>%
  mutate(age = as.factor(signif(age,1))) %>%
  ggplot(aes(x=exercise, y= cholesteral, color=age)) +
  geom_point() +
  geom_smooth(method='lm', se=F) +
  labs(x="Exercise [Hours per Week]", y="Cholesterol [Total cholesterol mg/dL]")
```

Understanding the full causal story is essential. Without an entire causal narrative, we might recommend inappropriate interventions; for example, a doctor might prescribe less exercise to reduce cholesterol in the case above. 

To deduce such causal stories, we need to apply the methodology of causal inference. 

